Chasing \rainbows and \oceanglints:\\ Inner working angle constraints for the Habitable Worlds Observatory

### 2023-05-20 src/tex/ms.tex
[showyourwork] overleaf sync

Liquid surface water is closely related to planetary habitability because it is essential for life as we know it.
%
To first order, its presence can be predicted by the flux a planet receives from its star.
%
The habitable zone (HZ) is defined as the range of semi-major axes around a given main sequence star, for which a rocky planet may have the right temperature to maintain liquid surface water \citep{kasting93}. 
%
However, there are many factors besides the incident stellar flux that influence a planet's ability to hold liquid surface water, such as the atmospheric pressure and geological activity. 
%
Detecting liquid surface water on an exoplanet would thus provide important context for the interpretation of possible biosignatures.
%
Variations of the flux and degree of polarization as an exoplanet orbits its host star---phase variations---could reveal scattering phenomena from condensed water \citep[e.g., rainbows;][]{bailey2007} or specular reflection, i.e. glint, from water oceans \citep[][]{2008Icar..195..927W}.
%
However, depending on a planetary system's architecture, such features can be obscured by the inner working angle (\IWA) of a telescope's coronagraph, shown schematically in Fig.~\ref{fig:annotated-orbit}.

@@

As shown in Fig.~\ref{fig:annotated-orbit}, we define the \emph{scattering angle} $\varphi$ as the planet--star--observer angle (large angles for back-scattering, small angles for forward-scattering) and the supplementary \emph{phase angle} $\alpha$ (zero at full phase and 180$^\circ$ at new phase). 
%
A planet on a perfectly face-on orbit is always seen at quadrature, $\varphi = \alpha = 90^\circ$, while a planet on an inclined orbit will exhibit a wider range of viewing angles; only perfectly edge-on orbits pass through all possible scattering angles. 
%
In some plots, we report \emph{degrees from quadrature}, so that both new and full phase are at \qty{90}{\degree} for edge-on orbits.

@@

Ocean glint is the specular (Fresnel) reflection of light off a smooth liquid surface. 
%
While glint is a familiar phenomenon on Earth, there is only one other body in the Solar System where it has been observed: the Cassini orbiter captured the glint of sunlight reflecting off the liquid hydrocarbon lakes near Titan's North Pole as it was observing Titan at crescent phases \citep{2010GeoRL..37.7104S}.
%
In the total flux, the glint signal is indeed expected to be most prominent at large planetary phase angles.
%
The maximum degree of polarization of the glint occurs at the Brewster angle, which is determined by the refractive indices at the interface; between air and liquid water this occurs at a reflection angle of \qty{53}{\degree} locally on the surface \citep[at a planetary phase angle of about \qty{106}%\qty{150}
{\degree}; see, e.g.,][]{Zugger_2010,treesandstam2019}.%{2008Icar..195..927W,Zugger_2010,treesandstam2019}
%
This angle therefore varies slightly with the wavelength and the composition of the ocean and the atmosphere.
%
The angular width of the glint in total and polarized flux generally broadens with increasing wave height \citep{2008Icar..195..927W, kopparla2018, treesandstam2019, trees2022}.\looseness=-1 \space In general, the glint on a planet will be strongest at red wavelengths because there Rayleigh scattering and attenuation in the atmosphere are smallest \citep{Zugger_2011}. 
%
This has been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}. 
%
At large phase angles, the forward scattering of light by cloud particles could be confused with glint \citep{Robinson_2010}.
%
This confusion can be resolved by measuring polarized phase curves at multiple wavelengths \citep{treesandstam2019}, as shown in \cref{fig:bottplot}, or by probing 
the rainbow, see below.

Glint provides evidence of specular reflecting surfaces, which, taken in the context of the temperature and types of condensates, may be interpreted as evidence for liquid water.
%
The presence of liquid water on the surface of a planet implies the presence of a water cycle which provides key context to its habitability and the interpretation of atmospheric biosignatures.
%
Aside from detecting ocean glint, the presence of surface water on Earth-like exoplanets can be established through less direct methods near quadrature such as spectropolarimetry, to infer ocean reflection \citep{trees2022}, or through rotational mapping, to identify bright continents and dark oceans \citep[e.g.,][]{2009ApJ...700..915C,lustig2019}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
    \centering
    \includegraphics{figures/figure-1-bott-plot.pdf}
    \script{create-figure-1-bott-plot.py}
    \caption{
        The phase curves of the normalized linearly polarized flux of a cloudy planet with a wavy ocean surface and an Earth-like atmosphere in an edge-on orbit; adapted from \citet{treesandstam2019}.  
        %
        The wind speed over the ocean, which determines the wave height \citep[][]{CoxMunk1954}, is \qty{7}{\meter\per\second}.
        %
        The clouds consist of spherical water droplets.
        %
        The cloud coverage fraction is 0.5 and while the cloud pattern is patchy, the glint is always cloud-free. 
        %
        The fluxes are normalized such that the total (unpolarized + polarized) flux equals the planet's geometric albedo at phase angle $\alpha = 0^\circ$ (scattering angle $\varphi = 180^\circ$).
        %
        Below the plot, we show the scattering angles of glories, rainbows, Rayleigh scattering, and ocean glint. Note that the features in the actual curves partly overlap with each other.
    }
    \label{fig:bottplot}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The National Academy of Sciences Astronomy \& Astrophysics 2020 Decadal Survey (Astro2020) \citep{decadal} recommended the first in a new \enquote{Great Observatories} program a telescope with the capability to detect signatures of habitability on about 25 HZ planets.
%
This requires an instrument with a coronagraph capable of high-contrast imaging and low-resolution spectroscopy at optical to near-infrared wavelengths. 
%
The precursor technology recommended by the survey lists \enquote{direct imaging to probe polarized ocean glint on terrestrial planets} as a priority capability \citep[Box E.1 in][]{decadal} for ground and space-based observatories.
%
Following the release of this report, NASA recently announced the start of the development of the Habitable Worlds Observatory (hereafter ``\HWO'').
%
The performance and characteristics (on- or off-axis, diameter, type of segmentation, polarimetry or photometry, and coronagraph, \IWA) of \HWO will significantly impact the expected exo-Earth yield \citep{Stark2019_exoplanetyield} and are still to be determined.
%
They will likely be heavily informed by the LUVOIR \citep{LUVOIR2019} and HabEx \citep{HabEx_2020} preparatory studies.
%
For generality, in this work, we report angular separations in milliarcseconds however, for ease of comparison, \cref{tab:IWA_OWA} lists the angular separations corresponding to multiples of $\lambda / D$ for a $\qty{6}{\meter}$ telescope, as suggested in Astro2020, at representative wavelengths $\lambda$ of $\qty{600}{\nano\meter}$ and $\qty{1}{\micro\meter}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table 1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
    \centering
    \caption{
        Inner and Outer Working Angles (\IWA and \OWA{}, respectively) in milli-arcseconds (\unit{\mas}) as functions of $\lambda / D$, at wavelengths $\lambda$ of \qty{600}{\nano\meter} and \qty{1}{\micro\meter}, assuming a telescope diameter $D$ of 6\,m.
    }
    \label{tab:IWA_OWA}
    \begin{tabular}{ c c c c } 
    \toprule
    & $\lambda/D$ & \unit{\mas} (at \qty{600}{\nano\meter}) & \unit{\mas} (at \qty{1}{\micro\meter}) \\
    \midrule
    \midrule
    IWA & 1 & 20.63 & 34.38 \\
    IWA & 2 & 41.25 & 68.75 \\
    IWA & 3 & 61.88 & 103.13 \\
    IWA & 4 & 82.51 & 137.51 \\
    \midrule
    OWA & 32 &  660.05 & 1100.08 \\
    OWA & 64 & 1320.09 & 2200.16 \\
    \bottomrule
    \end{tabular}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

A preliminary target list of nearby stars (maximum distance \qty{25}{\parsec}) around which \HWO may search for habitable rocky planets, assuming a \qty{6}{\meter} telescope, has been compiled: the \emph{NASA Exoplanet Exploration Program's Mission Star List for the Habitable Worlds Observatory}.%
\footnote{Available online: \url{https://exoplanets.nasa.gov/internal_resources/2645_NASA_ExEP_Target_List_HWO_Documentation_2023.pdf}}
%
This list consists of $\num{164}$ stars, the majority of which are Sun-like dwarfs: 66~F~dwarfs, 55~G~dwarfs, 40~K~dwarfs, and 3~M~dwarfs.
%
The target list is constructed assuming a maximum planet magnitude in the Cousins Rc band of $R_c = 31$ and a lower limit of \num{2.5e-11} for the planet-star flux ratio. 
% 
The adopted HZ limits are a semi-major axis of \qtyrange{0.95}{1.67}{\au} for a solar twin and planet sizes between \qtyrange{0.8}{1.4}{} Earth radii. 
%
For non-solar stars, the semi-major axis scales as square root of the bolometric luminosity normalized to the Sun. 
%
This range of orbital separation corresponds to the \enquote{conservative habitable zone} \citep{kasting93, kopparapu13}. 
%
In the remainder of this work, we consider hypothetical Earth-sized planets orbiting at the right distance from their parent star to receive the same flux as the Earth receives from the Sun.
%
By design, such planets should be observable by \HWO at quadrature phase.

@@

We first define the phase angle ranges over which the various phenomena --- glint, Rayleigh peak, rainbow and glory --- manifest themselves in \cref{tab:phase_ranges}, also shown in Fig.~\ref{fig:bottplot}. 
%
These ranges are all approximate, as the exact angles where these phenomena dominate depend on the details of the atmospheric and surface properties, as well as on the wavelengths being observed.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table 2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
    \centering
    \caption{
        The approximate ranges of the phase angle $\alpha$ for the different optical phenomena considered in this paper (cf. Fig.~\ref{fig:bottplot}).
    }
    \label{tab:phase_ranges}
    \begin{tabular}{ l c c c } 
        \toprule
        & $\alpha_\mathrm{min}$ & $\alpha_\mathrm{peak}$ & $\alpha_\mathrm{max}$ \\
        \midrule
        Glory       & \qty{0}{\degree}     & \qty{5}{\degree}     & \qty{10}{\degree} \\
        Rainbow    & \qty{22}{\degree}    & \qty{42}{\degree}    & \qty{63}{\degree} \\
        Rayleigh peak    & \qty{50}{\degree}    & \qty{70}{\degree}    & \qty{110}{\degree} \\
        Ocean Glint & \qty{130}{\degree}   & \qty{150}{\degree}   & \qty{170}{\degree} \\
        \bottomrule
    \end{tabular}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A coronagraphic mask can obscure part of a planet's orbit. As illustrated in \cref{fig:annotated-orbit} obscuration is a problem at small and large phase angles, when the planet is at small angular separation from its star.
The range of observable phase angles will depend on the \IWA of the coronagraph, as well as on the inclination and eccentricity of the exoplanet's orbit.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 3
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includestandalone{tikz/scattering-angle}
    \caption{
        The angular separation of a planet from its star as seen by the observer, $\delta$, is related to the scattering angle, $\varphi$. The star is at a distance $d_*$ from the observer, the planet orbits a distance $a$ from the star, and  the projected distance between the star and planet is  $r_\mathrm{proj}$. Not to scale. 
    }
    \label{fig:scattering-angle}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Most planets are not in edge-on orbits as seen from Earth.
For an orbit with an inclination angle $i$ ($i=0^\circ$ is face-on), we can identify the following two regimes:
\begin{enumerate}
    \item The coronagraph does not obscure any part of the sky-projected orbit: 
    the maximum phase angle coverage depends solely on $i$. 
        This holds for orbits that are relatively face-on (small $i$). 
    \item The coronagraph does obscure parts of the sky-projected orbit: the phase angle coverage depends only on the \IWA and the orbital semi-major axis $a$, following Eq.~\ref{eq:scattering_angle}. 
\end{enumerate}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 5
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}%[t]
    \centering
    \includegraphics{figures/figure-5-n-over-phase.pdf}
    \script{create-figure-5-n-over-phase.py}
    \caption{
        Cumulative distributions of the most extreme phase angles accessible for different \IWA and for randomly inclined, circular orbits (solid lines) and randomly inclined, eccentric orbits (dashed lines).
        %
        The top and bottom $x$-axes indicate the minimum and maximum accessible phase angles, respectively.
        %
        These angles are symmetric about quadrature (90$^\circ$).
        %
        The $y$-axis on the left indicates the number of planetary systems divided by the number of Monte Carlo samples, and is thus normalised to the number of systems on the target list.
        %
        The $y$-axis on the right indicates the number of HZ Earth-like planets that could be imaged at these phase angles, assuming an occurrence rate of $\eta_\oplus = \qty{24}{\percent}$.
        {\color{red}{the plot says 'elliptical orbits' instead of 'eccentric orbits'}}
    }
    \label{fig:betaallofit}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 7: Yarn ball
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
    \centering
    \includegraphics{figures/figure-7-yarnball.pdf}  
    \script{create-figure-7-yarnball.py}
    \caption{
        Examples of the eccentric orbits generated for the stellar sample. Each plot shows 10 examples of orbits randomly drawn for each star.
        The orbits are scaled to ensure an Earth-like incident flux. 
        %
        The concentric, dashed circles indicate inner working angles of \num{20.6}, \num{41.3}, \num{61.9}, and \qty{82.5}{\mas}.
        %, corresponding to 1, 2, 3, and 4\,$\lambda / D$, respectively (assuming $\lambda = \qty{600}{\nano\meter}$ and $D = \qty{6}{\meter}$).
        %
        Each system has been scaled such that its HZ has the same radius, hence the variation in the values of the \IWA{}s compared to the orbits.
        %
        It is clear that the \IWA can significantly affect the range of observable phases angles for each orbit.
    }
    \label{fig:ball-o-yarn}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

A second parameter that is crucial for detecting angular features like rainbows and ocean glint on Earth-like exoplanets is the contrast, i.e., the planet/star flux ratio. 
%
While not the main focus of this paper, we note that the total and polarized flux contrasts depend on the phase angle, due to a combination of the changing illumination fraction of the observable planetary disk and non-isotropic reflection by the planet.
%
We present some simplified examples in Fig.~\ref{fig:contrasts}. 
%
These calculations are by no means complete or exact, given the many variables that go into the modelling of reflected total and polarized fluxes of Earth-like exoplanets \citep{treesandstam2019,trees2022}.

>>>>

Liquid surface water is closely related to planetary habitability because it is essential for life as we know it.
%
To first order, its presence can be predicted by the flux a planet receives from its star.
%
The habitable zone (HZ) is defined as the range of semi-major axes around a given main sequence star, for which a rocky planet may have the right temperature to maintain liquid surface water \citep{kasting93}. 
%
However, there are many factors besides the incident stellar flux that influence a planet's ability to hold liquid surface water, such as the atmospheric pressure and geological activity. 
%
Detecting liquid surface water on an exoplanet would thus provide important context for the interpretation of possible biosignatures.
%
Variations of the flux and degree of polarization as an exoplanet orbits its host star---phase variations---could reveal scattering phenomena from condensed water \citep[e.g., rainbows;][]{bailey2007} or specular reflection, i.e. glint, from water oceans \citep[][]{2008Icar..195..927W}.
%
However, depending on a planetary system's architecture, such features can be obscured by the inner working angle (\IWA) of a telescope's coronagraph, shown schematically in \cref{fig:annotated-orbit}.

@@

As shown in \cref{fig:annotated-orbit}, we define the \emph{scattering angle} $\varphi$ as the planet--star--observer angle (large angles for back-scattering, small angles for forward-scattering) and the supplementary \emph{phase angle} $\alpha$ (\qty{0}{\degree} at full phase and \qty{180}{\degree} at new phase). 
%
A planet on a perfectly face-on orbit is always seen at quadrature, $\varphi = \alpha = \qty{90}{\degree}$, while a planet on an inclined orbit will exhibit a wider range of viewing angles; only perfectly edge-on orbits pass through all possible scattering angles. 
%
In some plots, we report \emph{degrees from quadrature}, so that both new and full phase are at \qty{90}{\degree} for edge-on orbits.

@@

Ocean glint is the specular (Fresnel) reflection of light off a smooth liquid surface. 
%
While glint is a familiar phenomenon on Earth, there is only one other body in the Solar System where it has been observed: the Cassini orbiter captured the glint of sunlight reflecting off the liquid hydrocarbon lakes near Titan's North Pole as it was observing Titan at crescent phases \citep{2010GeoRL..37.7104S}.
%
In the total flux, the glint signal is indeed expected to be most prominent at large planetary phase angles.
%
The maximum degree of polarization of the glint occurs at the Brewster angle, which is determined by the refractive indices at the interface; between air and liquid water this occurs at a reflection angle of \qty{53}{\degree} locally on the surface \citep[at a planetary phase angle of about \qty{106}%\qty{150}
{\degree}; see, e.g.,][]{Zugger_2010,treesandstam2019}.%{2008Icar..195..927W,Zugger_2010,treesandstam2019}
%
This angle therefore varies slightly with the wavelength and the composition of the ocean and the atmosphere.
%
The angular width of the glint in total and polarized flux generally broadens with increasing wave height \citep{2008Icar..195..927W, kopparla2018, treesandstam2019, trees2022}.
In general, the glint on a planet will be strongest at red wavelengths because there, Rayleigh scattering and attenuation in the atmosphere are smallest \citep{Zugger_2011}. 
%
This has been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}. 
%
At large phase angles, the forward scattering of light by cloud particles could be confused with glint \citep{Robinson_2010}.
%
This confusion can be resolved by measuring polarized phase curves at multiple wavelengths \citep{treesandstam2019}, as shown in \cref{fig:bottplot}, or by probing 
the rainbow, see below.

Glint provides evidence of specular reflecting surfaces, which, taken in the context of the temperature and types of condensates, may be interpreted as evidence for liquid water.
%
The presence of liquid water on the surface of a planet implies the presence of a water cycle, which provides key context to its habitability and the interpretation of atmospheric biosignatures.
%
Aside from detecting ocean glint, the presence of surface water on Earth-like exoplanets can be established through less direct methods near quadrature such as spectropolarimetry, to infer ocean reflection \citep{trees2022}, or through rotational mapping, to identify bright continents and dark oceans \citep[e.g.,][]{2009ApJ...700..915C,lustig2019}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
    \centering
    \includegraphics{figures/figure-1-bott-plot.pdf}
    \script{create-figure-1-bott-plot.py}
    \caption{
        The phase curves of the normalized linearly polarized flux of a cloudy planet with a wavy ocean surface and an Earth-like atmosphere in an edge-on orbit; adapted from \citet{treesandstam2019}.  
        %
        The wind speed over the ocean, which determines the wave height \citep[][]{CoxMunk1954}, is \qty{7}{\meter\per\second}.
        %
        The clouds consist of spherical water droplets.
        %
        The cloud coverage fraction is 0.5 and while the cloud pattern is patchy, the glint is always cloud-free. 
        %
        The fluxes are normalized such that the total (unpolarized + polarized) flux equals the planet's geometric albedo at phase angle $\alpha = \qty{0}{\degree}$ (scattering angle $\varphi = \qty{180}{\degree}$).
        %
        Below the plot, we show the scattering angles of glories, rainbows, Rayleigh scattering, and ocean glint. Note that the features in the actual curves partly overlap with each other.
    }
    \label{fig:bottplot}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The National Academy of Sciences Astronomy \& Astrophysics 2020 Decadal Survey (Astro2020) \citep{decadal} recommended the first in a new \enquote{Great Observatories} program a telescope with the capability to detect signatures of habitability on about 25 HZ planets.
%
This requires an instrument with a coronagraph capable of high-contrast imaging and low-resolution spectroscopy at optical to near-infrared wavelengths. 
%
The precursor technology recommended by the survey lists \enquote{direct imaging to probe polarized ocean glint on terrestrial planets} as a priority capability \citep[Box E.1 in][]{decadal} for ground and space-based observatories.
%
Following the release of this report, NASA recently announced the start of the development of the Habitable Worlds Observatory (hereafter \enquote{\HWO}).
%
The performance and characteristics (on- or off-axis, diameter, type of segmentation, polarimetry or photometry, and coronagraph, \IWA) of \HWO will significantly impact the expected exo-Earth yield \citep{Stark2019_exoplanetyield} and are still to be determined.
%
They will likely be heavily informed by the LUVOIR \citep{LUVOIR2019} and HabEx \citep{HabEx_2020} preparatory studies.
%
For generality, in this work, we report angular separations in milliarcseconds; however, for ease of comparison, \cref{tab:IWA_OWA} lists the angular separations corresponding to multiples of $\lambda / D$ for a $\qty{6}{\meter}$ telescope, as suggested in Astro2020, at representative wavelengths $\lambda$ of $\qty{600}{\nano\meter}$ and $\qty{1}{\micro\meter}$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table 1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
    \centering
    \caption{
        Inner and Outer Working Angles (\IWA and \OWA{}, respectively) in milliarcseconds (\unit{\mas}) as functions of $\lambda / D$, at wavelengths $\lambda$ of \qty{600}{\nano\meter} and \qty{1}{\micro\meter}, assuming a telescope diameter $D$ of 6\,m.
    }
    \label{tab:IWA_OWA}
    \begin{tabular}{ c c c c } 
    \toprule
    & $\lambda/D$ & \unit{\mas} (at \qty{600}{\nano\meter}) & \unit{\mas} (at \qty{1}{\micro\meter}) \\
    \midrule
    \midrule
    IWA & 1 & 20.63 & 34.38 \\
    IWA & 2 & 41.25 & 68.75 \\
    IWA & 3 & 61.88 & 103.13 \\
    IWA & 4 & 82.51 & 137.51 \\
    \midrule
    OWA & 32 &  660.05 & 1100.08 \\
    OWA & 64 & 1320.09 & 2200.16 \\
    \bottomrule
    \end{tabular}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

A preliminary target list of nearby stars (maximum distance \qty{25}{\parsec}) around which \HWO may search for habitable rocky planets, assuming a \qty{6}{\meter} telescope, has been compiled: the \emph{NASA Exoplanet Exploration Program's Mission Star List for the Habitable Worlds Observatory}.%
\footnote{Available online: \url{https://exoplanets.nasa.gov/internal_resources/2645_NASA_ExEP_Target_List_HWO_Documentation_2023.pdf}}
%
This list consists of $\num{164}$ stars, the majority of which are Sun-like dwarfs: 66~F~dwarfs, 55~G~dwarfs, 40~K~dwarfs, and 3~M~dwarfs.
%
The target list is constructed assuming a maximum planet magnitude in the Cousins Rc band of $R_c = 31$ and a lower limit of \num{2.5e-11} for the planet-star flux ratio. 
% 
The adopted HZ limits are a semi-major axis of \qtyrange{0.95}{1.67}{\au} for a solar twin and planet sizes between \qtyrange{0.8}{1.4}{} Earth radii. 
%
For non-solar stars, the semi-major axis scales as the square root of the bolometric luminosity normalized to the Sun. 
%
This range of orbital separation corresponds to the \enquote{conservative habitable zone} \citep{kasting93, kopparapu13}. 
%
In the remainder of this work, we consider hypothetical Earth-sized planets orbiting at the right distance from their parent star to receive the same flux as the Earth receives from the Sun.
%
By design, such planets should be observable by \HWO at quadrature phase.

@@

We first define the phase angle ranges over which the various phenomena---glint, Rayleigh peak, rainbow, and glory---manifest themselves in \cref{tab:phase_ranges}, also shown in \cref{fig:bottplot}. 
%
These ranges are all approximate, as the exact angles where these phenomena dominate depend on the details of the atmospheric and surface properties, as well as on the wavelengths being observed.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table 2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
    \centering
    \caption{
        The approximate ranges of the phase angle $\alpha$ for the different optical phenomena considered in this paper (cf. \cref{fig:bottplot}).
    }
    \label{tab:phase_ranges}
    \begin{tabular}{ l c c c } 
        \toprule
        & $\alpha_\mathrm{min}$ & $\alpha_\mathrm{peak}$ & $\alpha_\mathrm{max}$ \\
        \midrule
        Glory       & \qty{0}{\degree}     & \qty{5}{\degree}     & \qty{10}{\degree} \\
        Rainbow    & \qty{22}{\degree}    & \qty{42}{\degree}    & \qty{63}{\degree} \\
        Rayleigh peak    & \qty{50}{\degree}    & \qty{70}{\degree}    & \qty{110}{\degree} \\
        Ocean Glint & \qty{130}{\degree}   & \qty{150}{\degree}   & \qty{170}{\degree} \\
        \bottomrule
    \end{tabular}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A coronagraphic mask can obscure part of a planet's orbit. 
As illustrated in \cref{fig:annotated-orbit}, obscuration is a problem at small and large phase angles, when the planet is at small angular separation from its star.
The range of observable phase angles will depend on the \IWA of the coronagraph, as well as on the inclination and eccentricity of the exoplanet's orbit.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 3
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
    \centering
    \includestandalone{tikz/scattering-angle}
    \caption{
        The angular separation of a planet from its star as seen by the observer, $\delta$, is related to the scattering angle $\varphi$. 
        The star is at a distance $d_*$ from the observer, the planet orbits a distance $a$ from the star, and  the projected distance between the star and planet is  $r_\mathrm{proj}$. 
        Not to scale. 
    }
    \label{fig:scattering-angle}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Most planets are not in edge-on orbits as seen from Earth.
For an orbit with an inclination angle $i$ ($i=\qty{0}{\degree}$ is face-on), we can identify the following two regimes:
\begin{enumerate}
    \item The coronagraph does not obscure any part of the sky-projected orbit: 
    the maximum phase angle coverage depends solely on $i$. 
        This holds for orbits that are relatively face-on (small $i$). 
    \item The coronagraph does obscure parts of the sky-projected orbit: the phase angle coverage depends only on the \IWA and the orbital semi-major axis $a$, following \cref{eq:scattering_angle}. 
\end{enumerate}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 5
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}%[t]
    \centering
    \includegraphics{figures/figure-5-n-over-phase.pdf}
    \script{create-figure-5-n-over-phase.py}
    \caption{
        Cumulative distributions of the most extreme phase angles accessible for different \IWA and for randomly inclined, circular orbits (solid lines) and randomly inclined, eccentric orbits (dashed lines).
        %
        The top and bottom $x$-axes indicate the minimum and maximum accessible phase angles, respectively.
        %
        These angles are symmetric about quadrature (\qty{90}{\degree}).
        %
        The $y$-axis on the left indicates the number of planetary systems divided by the number of Monte Carlo samples, and is thus normalised to the number of systems on the target list.
        %
        The $y$-axis on the right indicates the number of HZ Earth-like planets that could be imaged at these phase angles, assuming an occurrence rate of $\eta_\oplus = \qty{24}{\percent}$.
        {\color{red}{the plot says 'elliptical orbits' instead of 'eccentric orbits'}}
    }
    \label{fig:betaallofit}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Figure 7: Yarn ball
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
    \centering
    \includegraphics{figures/figure-7-yarnball.pdf}  
    \script{create-figure-7-yarnball.py}
    \caption{
        Examples of the eccentric orbits generated for the stellar sample. 
        Each plot shows 10 examples of orbits randomly drawn for each star.
        The orbits are scaled to ensure an Earth-like incident flux. 
        The concentric, dashed circles indicate inner working angles of \num{20.6}, \num{41.3}, \num{61.9}, and \qty{82.5}{\mas}.
        %, corresponding to 1, 2, 3, and 4\,$\lambda / D$, respectively (assuming $\lambda = \qty{600}{\nano\meter}$ and $D = \qty{6}{\meter}$).
        %
        Each system has been scaled such that its HZ has the same radius, hence the variation in the values of the \IWA{}s compared to the orbits.
        It is clear that the \IWA can significantly affect the range of observable phases angles for each orbit.
    }
    \label{fig:ball-o-yarn}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

A second parameter that is crucial for detecting angular features like rainbows and ocean glint on Earth-like exoplanets is the contrast, that is, the planet/star flux ratio. 
%
While not the main focus of this paper, we note that the total and polarized flux contrasts depend on the phase angle, due to a combination of the changing illumination fraction of the observable planetary disk and non-isotropic reflection by the planet.
%
We present some simplified examples in \cref{fig:contrasts}. 
%
These calculations are by no means complete or exact, given the many variables that go into the modelling of reflected total and polarized fluxes of Earth-like exoplanets \citep{treesandstam2019,trees2022}.

The NASA Exoplanet Exploration Program’s Mission Star List for \HWO includes for every star an estimated contrast for an Earth-twin in the habitable zone.
%
The contrast is given by \citep[see, e.g.,][]{2023A&A...671A.165M}
\begin{equation}
    C(\alpha) 
    = F_p(\alpha) / F_* 
    = A_\mathrm{G} \cdot \phi(\alpha) \cdot \left( R_p / a(\alpha) \right)^2 \,,
    \label{eq:contrast}
\end{equation}
where $A_\mathrm{G}$ is the planet's geometric albedo, $R_p$ its radius, $\phi(\alpha)$ is the phase function at phase angle $\alpha$ (normalized to \num{1.0} at full phase, thus at $\alpha=\qty{0}{\degree}$), and $a$ is the distance between the planet and its host star.
%
The assumptions that went into the contrast estimations for \HWO are that $a$ is constant (the orbits are circular), the geometric albedo $A_\mathrm{G}$ is \num{0.2}, and that the phase function $\phi(\alpha)$ can be described by a Lambertian phase function \citep{Russell1916}, which assumes isotropic reflection by the planet. 
%
We assume that the calculated contrasts in the list correspond to the reflected flux at quadrature and use \cref{eq:contrast} to calculate the reflected light fraction at $\alpha=\qty{90}{\degree}$ as $C(90) = \qty{6.37}{\percent}$.

We use the phase functions at $\qty{670}{\nano\meter}$ from \citet{treesandstam2019} for an Earth-like planet with an ocean surface, a wind-speed of $\qty{7}{\meter\per\second}$, and patchy liquid water clouds. 
%
We correct the Star List contrast by dividing by the $6.37\%$ reflected light fraction of the Lambertian phase function and multiplying by the reflected light fraction of the \citet{treesandstam2019} model of $5.64\%$.
%
We map the orbital phase for circular orbits to the phase angle and on-sky separation, given an inclination $i$ of \qty{90}{\degree}. 
%
Together, they give the reflected light contrast as a function of separation for a full orbit.
%
We multiply this flux with the degree of polarization to obtain the polarized contrast.
%
\cref{fig:contrasts} shows the resulting contrast curves for three of the closest target stars: the M~dwarf Lalande~21185, the K~dwarf $\epsilon$~Eri, and the G~dwarf $\alpha$~Cen~A
